---
title: "Checking habitat data exported from TransectMeasure"
author: "Claude Spencer & Brooke Gibbons"
date: "2023-11-13"
output:
  html_vignette:
    toc: true
    code_folding: hide
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Checking habitat data exported from TransectMeasure}
  %\VignetteEncoding{UTF-8}
---

This script documents the necessary steps to take raw data exported from TransectMeasure, check for any errors in the annotation and format it into tidy data. You are required to also have a GlobalArchive format metadata file "*_Metadata.csv".

# R set up
Load the necessary libraries.

```{r libraries, results='hide', warning=FALSE, message=FALSE}
# install.packages('remotes')
library('remotes')
options(timeout=9999999)
# remotes::install_github("GlobalArchiveManual/CheckEM")
library(CheckEM)
library(tidyverse)
library(ggplot2)
library(ggbeeswarm)
library(leaflet)
library(leaflet.minicharts)
library(RColorBrewer)
library(here)
library(tidyverse)
```

Set the study name.

```{r name}
name <- "example-bruv-workflow"
```

# Read in the data
Load and format the metadata.

```{r metadata}
metadata <- read_metadata(here::here("r-workflows/data/raw/")) %>%
    dplyr::select(campaignid, sample, longitude_dd, latitude_dd, date_time, location, site, depth_m, successful_count, successful_length, successful_habitat_forward, successful_habitat_backward) %>%
  glimpse()
```

Save the metadata as an R data file (this creates a lighter file than saving as a .csv or similar).

```{r save-metadata}
saveRDS(metadata, file = here::here(paste0("r-workflows/data/tidy/", name, "_metadata.rds")))
```

Read in the points data exported from Transect Measure, and then filter this into 2 dataframes for relief and habitat annotations.

```{r points}
points <- read_TM(here::here("r-workflows/data/raw/"),
                               sample = "opcode")
```

Filter the data to only include habitat annotations.

```{r habitat}
habitat <- points %>%
  dplyr::filter(relief_annotated %in% "no") %>%
  dplyr::select(campaignid, sample, starts_with("level"), scientific) %>%
  glimpse()
```

Filter the data to only include relief annotations.

```{r relief}
relief <- points %>%
  dplyr::filter(relief_annotated %in% "yes") %>%
  dplyr::select(campaignid, sample, starts_with("level"), scientific) %>%
  glimpse()
```

# Check the data for a variety of errors
Set the number of expected annotations to check against.

```{r set-points}
num.points <- 20
```

Check if there are habitat annotations with an unexpected number of annotation points.

```{r wrong-habitat}
wrong.points.habitat <- habitat %>%
  group_by(campaignid, sample) %>%
  summarise(points.annotated = n()) %>%
  left_join(metadata) %>%
  dplyr::mutate(expected = case_when(successful_habitat_forward %in% "Yes" & successful_habitat_backward %in% "Yes" ~ num.points * 2, successful_habitat_forward %in% "Yes" & successful_habitat_backward %in% "No" ~ num.points * 1, successful_habitat_forward %in% "No" & successful_habitat_backward %in% "Yes" ~ num.points * 1, successful_habitat_forward %in% "No" & successful_habitat_backward %in% "No" ~ num.points * 0)) %>%
  dplyr::filter(!points.annotated == expected) %>%
  glimpse()
```

Check if there are habitat annotations with an unexpected number of annotation points.

```{r wrong-relief}
wrong.points.relief <- relief %>%
  group_by(campaignid, sample) %>%
  summarise(points.annotated = n()) %>%
  left_join(metadata) %>%
  dplyr::mutate(expected = case_when(successful_habitat_forward %in% "Yes" & successful_habitat_backward %in% "Yes" ~ num.points * 2, successful_habitat_forward %in% "Yes" & successful_habitat_backward %in% "No" ~ num.points * 1, successful_habitat_forward %in% "No" & successful_habitat_backward %in% "Yes" ~ num.points * 1, successful_habitat_forward %in% "No" & successful_habitat_backward %in% "No" ~ num.points * 0)) %>%
  dplyr::filter(!points.annotated == expected) %>%
  glimpse()
```

Check to see if there are any samples in habitat that don't have a match in the metadata.

```{r missing-metadata}
habitat.missing.metadata <- anti_join(habitat, metadata, by = c("campaignid", "sample")) %>%
  glimpse()
```

Check to see if there are any samples in the metadata that don't have matching habitat.

```{r missing-habitat}
metadata.missing.habitat <- anti_join(metadata, habitat, by = c("campaignid", "sample")) %>%
  glimpse()
```

# Format and save the final tidy dataset
Make a vector of columns included in the schema file to ensure proper joining with your tidy habitat dataset.

```{r add-columns}
catami_cols <- c("level_1" = NA,
                 "level_2" = NA,
                 "level_3" = NA,
                 "level_4" = NA,
                 "level_5" = NA,
                 "level_6" = NA,
                 "level_7" = NA,
                 "level_8" = NA,
                 "family" = NA,
                 "genus" = NA,
                 "species" = NA)

```

Tidy the habitat data by joining with the complete schema file that is loaded with the CheckEM package.


```{r tidy-habitat}
tidy.habitat <- habitat %>%
  dplyr::mutate(count = 1) %>%
  add_column(!!!catami_cols[!names(catami_cols) %in% names(.)]) %>%
  left_join(catami) %>%
  dplyr::select(campaignid, sample, count, starts_with("level"), family, genus, species) %>%
  dplyr::filter(!level_2 %in% c("","Unscorable", NA)) %>%
  group_by(campaignid, sample, across(starts_with("level")), family, genus, species) %>%
  dplyr::summarise(count = sum(count)) %>%
  ungroup() %>%
  dplyr::select(campaignid, sample, level_1, everything()) %>%
  glimpse()
```

Save the tidy habitat data as an R data file.

```{r save-tidy-habitat}
saveRDS(tidy.habitat, file = here::here(paste0("r-workflows/data/staging/", name, "_habitat.rds")))
```

Tidy the relief data by joining with the complete schema file that is loaded with the CheckEM package.

```{r tidy-relef}
tidy.relief <- relief %>%
  dplyr::select(campaignid, sample, starts_with("level"), scientific) %>%
  dplyr::filter(!level_2 %in% c("","Unscorable", NA)) %>%
  dplyr::mutate(count = 1) %>%
  add_column(!!!catami_cols[!names(catami_cols) %in% names(.)]) %>%
  left_join(catami) %>%
  group_by(campaignid, sample, across(starts_with("level"))) %>%
  dplyr::summarise(count = sum(count)) %>%
  ungroup() %>%
  glimpse()
```

Save the tidy relief data as an R data file.

```{r save-tidy-relief}
saveRDS(tidy.relief, file = here::here(paste0("r-workflows/data/staging/", name, "_relief.rds")))
```