---
title: "Generate Generalized Additive Model predictions of fish"
author: "Claude Spencer & Brooke Gibbons"
date: "2023-11-13"
output:
  html_vignette:
    toc: true
    code_folding: hide
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Generate Generalized Additive Model predictions of fish}
  %\VignetteEncoding{UTF-8}
---
This script takes the tidy datasets created in previous steps, along with the top models selection using FSSgam, and creates GAM regression plots. These plots are useful to correctly interpret the relationships between response and predictor variables, and allow for ecological investigation of top models.

```{r install-fssgam, include=FALSE}
devtools::install_github("beckyfisher/FSSgam_package", upgrade = "never")
```

# R setup
Load libraries. All packages are available through CRAN, aside from ‘FSSgam’, which can be installed following the instructions provided in the GitHub repository https://github.com/beckyfisher/FSSgam.

```{r libraries, results='hide', warning=FALSE, message=FALSE}
library(CheckEM)
library(tidyverse)
library(mgcv)
library(devtools)
library(FSSgam)
library(here)
library(ggplot2)
library(ggnewscale)
library(viridis)
library(terra)
library(sf)
library(patchwork)
```

Set the study name. Study names are used throughout to make for reproducible code that can be easily shifted between different campaigns and projects while still providing meaningful filenames.

```{r study-name}
name <- 'example-bruv-workflow'
```

Load the count data (MaxN). This data is crated in the previous workflow, 'Format and visualise fish data'.

```{r load-count}
tidy.count <- readRDS(here::here(paste0('r-workflows/data/tidy/', 
                      name,'_tidy-count.rds'))) %>%
  dplyr::filter(!is.na(mbdepth)) %>% # If this filters out any then you need to go back and create your spatial layers
  glimpse()
```

Load the fish length data. This data is created in the previous workflow, 'Format & visualise fish data'.

```{r read-length}
tidy.length <- readRDS(here::here(paste0('r-workflows/data/tidy/', 
                      name,'_tidy-length.rds'))) %>%
  dplyr::filter(!is.na(mbdepth)) %>% # If this filters out any then you need to go back and create your spatial layers
  glimpse()
```

Combine the count and length datasets to make for more simple code.

```{r combine-data}
dat <- bind_rows(tidy.count, tidy.length)
```

Filter the data to only include 1 response variable. The example included here models fish total abundance, and uses a dummy model not selected in FSSgam, in order to provide all the necessary code to produce fish predictions for all types of response variable.

```{r filter-data}
# MODEL Total abundance (depth + status) ----
dat_total <- dat %>% 
  dplyr::filter(response %in% "total_abundance") %>%
  dplyr::mutate(status = rep(c("Fished", "No-take"), times = 16))
```

Manually set the top model selected in the previous workflow, 'Select top model for fish using FSSgam'. Take care to make sure that it matches the model provided by FSSgam (eg. number of knots (k), smooth terms (bs), random effects). Remember this is an example model, hence the warning message if using these example data!

```{r set-model}
mod <- gam(number ~ s(detrended, k = 3, bs = 'cr') + status + s(aspect, k = 3, bs = 'cc'), family = tw(), data = dat_total)
```

Generate a dataframe with the values for detrended bathymetry while holding the other predictor variables at their mean.

```{r expand-detrended}
testdata <- expand.grid(detrended = seq(min(dat$detrended), max(dat$detrended),length.out = 20),
                        aspect = mean(mod$model$aspect),
                        status = c("Fished", "No-take")) %>%
  distinct() %>%
  glimpse()
```

Predict total abundance across the range of detrended bathymetry, using the 'predict.gam' function. 

```{r predict-detrended}
fits <- predict.gam(mod, newdata = testdata, type = 'response', se.fit = T)
```

Combine the two dataframes to create mean and standard deviation values for the relationship between total abundance and detrended bathymetry.

```{r mean-detrended}
predicts_total_detrended <- testdata %>%
  data.frame(fits) %>%
  group_by(detrended) %>% # Only change here
  summarise(number = mean(fit), se.fit = mean(se.fit)) %>%
  ungroup()
```

Generate a dataframe with the values for status (factor variable) while holding the other predictor variables at their mean.

```{r expand-status}
testdata <- expand.grid(status = c("Fished", "No-take"),
                        detrended = mean(mod$model$detrended),
                        aspect = mean(mod$model$aspect)) %>%
  distinct() %>%
  glimpse()
```

Predict total abundance for the two levels of the factor variable status, using the 'predict.gam' function. 

```{r predict-status}
fits <- predict.gam(mod, newdata = testdata, type = 'response', se.fit = T)
```

Combine the two dataframes to create mean and standard deviation values for the relationship between total abundance and status.

```{r mean-status}
predicts_total_status <- testdata %>%
  data.frame(fits) %>%
  group_by(status) %>% # Only change here
  summarise(number = mean(fit), se.fit = mean(se.fit)) %>%
  ungroup()
```

Generate a dataframe with the values for aspect while holding the other predictor variables at their mean.

```{r expand-aspect}
testdata <- expand.grid(aspect = seq(min(dat$aspect), max(dat$aspect),length.out = 20),
                        status = c("Fished", "No-take"),
                        detrended = mean(mod$model$detrended)) %>%
  distinct() %>%
  glimpse()
```

Predict total abundance across the range of aspect, using the 'predict.gam' function. 

```{r predict-aspect}
fits <- predict.gam(mod, newdata = testdata, type = 'response', se.fit = T)
```

Combine the two dataframes to create mean and standard deviation values for the relationship between total abundance and aspect.

```{r mean-aspect}
predicts_total_aspect <- testdata %>%
  data.frame(fits) %>%
  group_by(aspect) %>% # Only change here
  summarise(number = mean(fit), se.fit = mean(se.fit))%>%
  ungroup()
```

Create tidy plots to visualise the relationship between fish total abundance and detrended bathymetry.

```{r plot-detrended}
gg_total_detrended <- ggplot() +
  geom_point(data = dat_total, aes(x = detrended, y = number),  alpha = 0.2, size = 1, show.legend = F) +
  geom_line(data = predicts_total_detrended, aes(x = detrended, y = number), alpha = 0.5) +
  geom_line(data = predicts_total_detrended, aes(x = detrended, y = number - se.fit), linetype = "dashed", alpha = 0.5) +
  geom_line(data = predicts_total_detrended, aes(x = detrended, y = number + se.fit), linetype = "dashed", alpha = 0.5) +
  theme_classic() +
  labs(x = "Detrended bathymetry", y = "", title = "Total abundance") +
  theme(plot.title = element_text(hjust = 0))
gg_total_detrended
```

Create tidy plots to visualise the relationship between fish total abundance and status. Note that as this plot is for a factor variable, the relationship is represented as bar plots.

```{r plot-status}
gg_total_status <- ggplot() +
  geom_bar(data = predicts_total_status, aes(x = status, y = number, fill = status), stat = "identity", show.legend = F) +
  scale_fill_manual(values = c("No-take" = "#7bbc63", 
                                 "Fished" = "#b9e6fb")) +
  geom_errorbar(data = predicts_total_status, aes(x = status, ymin = number - se.fit, ymax = number + se.fit), width = 0.5) +
  theme_classic() +
  labs(x = "Status", y = "")
gg_total_status
```

Create tidy plots to visualise the relationship between fish total abundance and aspect. Note that aspect is a circular variable (compasss bearing from 0 to 360 degrees), and thus the relationship is represented as a polar plot.

```{r plot-aspect}
gg_total_aspect <- ggplot() +
  geom_point(data = dat_total, aes(x = aspect, y = number),  alpha = 0.2, size = 1, show.legend = F) +
  geom_line(data = predicts_total_aspect, aes(x = aspect, y = number), alpha = 0.5) +
  geom_line(data = predicts_total_aspect, aes(x = aspect, y = number - se.fit), linetype = "dashed", alpha = 0.5) +
  geom_line(data = predicts_total_aspect, aes(x = aspect, y = number + se.fit), linetype = "dashed", alpha = 0.5) +
  theme_classic() +
  labs(x = "Aspect", y = "", title = "Total abundance") +
  theme(plot.title = element_text(hjust = 0)) +
  coord_polar()
gg_total_aspect
```