---
title: "Format & visualise fish data"
author: "Claude Spencer & Brooke Gibbons"
date: "2023-11-13"
output:
  html_vignette:
    toc: true
    code_folding: hide
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Format & visualise fish data}
  %\VignetteEncoding{UTF-8}
---

This script takes the checked fish data from the previous workflow steps, visualises the data and exports it into a format suitable for modelling. The exploratory visualisation of the data allows for trends and patterns in the raw data to be investigated.

# R set up
Load the necessary libraries.

```{r libraries, results='hide', warning=FALSE, message=FALSE}
library('remotes')
options(timeout=9999999)
# remotes::install_github("GlobalArchiveManual/CheckEM")
library(CheckEM)
library(tidyverse)
library(ggplot2)
library(sf)
library(here)
library(leaflet)
```

Set the study name.

```{r study-name}
name <- "example-bruv-workflow"
```

# Read in the data
Load and format the metadata.

```{r load-metadata}
metadata.bathy.derivatives <- readRDS(here::here(paste0("r-workflows/data/tidy/", name, "_metadata-bathymetry-derivatives.rds"))) %>%
  glimpse()
```

Load and format the habitat data. Here we create a 'reef' column that consists of reef-forming habitat found in the study area, which will be used in subsequent fish modelling scripts.

```{r load-habitat}
habitat <- readRDS(here::here(paste0("r-workflows/data/tidy/", name, "_tidy-habitat.rds"))) %>%
  dplyr::mutate(number = count / total_points_annotated) %>%
  dplyr::select(campaignid, sample, longitude_dd, latitude_dd, depth, mbdepth, slope, aspect, tpi, tri, roughness, detrended, habitat, number, mean_relief, sd_relief) %>%
  pivot_wider(names_from = habitat, values_from = number, values_fill = 0) %>%
  CheckEM::clean_names() %>%
  dplyr::mutate(reef = consolidated_hard + macroalgae + seagrasses + sessile_invertebrates) %>%           
  dplyr::select(-c(consolidated_hard, macroalgae, seagrasses, sessile_invertebrates, unconsolidated_soft)) %>%
  glimpse()
```

Load length of maturity data. The maturity dataset is loaded when the package is installed.

```{r load-maturity}
maturity.mean <- maturity %>%
  dplyr::group_by(family, genus, species, sex) %>%
  dplyr::slice(which.min(l50_mm)) %>%
  ungroup() %>%
  dplyr::group_by(family, genus, species) %>%
  dplyr::summarise(l50 = mean(l50_mm)) %>%
  ungroup() %>%
  glimpse()
```

Load and format count data, and transform this into total abundance and species richness. These two response variables will be used in subsequent fish modelling scripts.

```{r tidy-count}
tidy.count <- readRDS(here::here(paste0("r-workflows/data/staging/", name, "_complete-count.rds"))) %>%
  dplyr::filter(campaignid %in% "2023-03_SwC_stereo-BRUVs") %>% # Remove Pt Cloates campaign for now             
  dplyr::group_by(campaignid, sample, status, scientific) %>%
  dplyr::summarise(count = sum(count)) %>%
  pivot_wider(names_from = "scientific", values_from = count, values_fill = 0) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(total_abundance = rowSums(.[, 4:(ncol(.))], na.rm = TRUE), species_richness = rowSums(.[, 4:(ncol(.))] > 0)) %>% 
  dplyr::select(campaignid, sample, status, total_abundance, species_richness) %>%
  pivot_longer(cols = c("total_abundance", "species_richness"), names_to = "response", values_to = "number") %>%
  dplyr::left_join(habitat) %>%
  glimpse()
```

Load and format length data.

```{r load-lengths}
lengths <- readRDS(here::here(paste0("r-workflows/data/staging/", name, "_complete-length.rds"))) %>%
  dplyr::filter(campaignid %in% "2023-03_SwC_stereo-BRUVs") %>% # Remove Pt Cloates campaign for now 
  # dplyr::mutate(depth = as.numeric(depth)) %>%                      
  left_join(habitat) %>%
  left_join(maturity.mean) %>%
  dplyr::mutate(number = 1) %>%
  glimpse()
```

# Format data for visualisation and modelling
Filter length data to only include 'indicator species'. In our case study, these are three species that have been defined as bioregional-scale demersal fish indicators by the West Australian Department of Primary Industries and Regional Development. If there is no equivalent in your study area, we suggest using ecologically-relevant and highly-targeted species.

```{r indicator species}
indicator.species <- lengths %>%
  dplyr::mutate(scientific = paste(genus, species, sep = " ")) %>%
  dplyr::filter(scientific %in% c("Choerodon rubescens", "Chrysophrys auratus", "Glaucosoma hebraicum")) %>%
  glimpse()
```

Create a unique list of the samples per campaign, to fill the zeroes back into processed length data.

```{r metadata-length}
metadata.length <- lengths %>%
  distinct(campaignid, sample, status) %>%
  glimpse()
```

Create a dataframe of the indicator species greater than length of maturity.

```{r greater-than-maturity}
greater.mat <- indicator.species %>%
  dplyr::filter(length_mm > l50) %>%
  dplyr::group_by(campaignid, sample) %>%
  dplyr::summarise(number = sum(number)) %>%
  ungroup() %>%
  right_join(metadata.length) %>%
  dplyr::mutate(number = ifelse(is.na(number), 0, number)) %>%
  dplyr::mutate(response = "greater than Lm") %>%
  left_join(habitat) %>%
  dplyr::glimpse()
```

Create a dataframe of the indicator species smaller than length of maturity.

```{r smaller-than-maturity}
smaller.mat <- indicator.species %>%
  dplyr::filter(length_mm < l50) %>%
  dplyr::group_by(campaignid, sample) %>%
  dplyr::summarise(number = sum(number)) %>%
  ungroup() %>%
  right_join(metadata.length) %>%
  dplyr::mutate(number = ifelse(is.na(number), 0, number)) %>%
  dplyr::mutate(response = "smaller than Lm") %>%
  left_join(habitat) %>%
  dplyr::glimpse()
```

Join the two datasets.

```{r tidy-length}
tidy.length <- bind_rows(greater.mat, smaller.mat) %>%
  glimpse()
```

# Visualise the fish count data
Filter the count (maxn) data to only include one response. In this example the count metrics included are total abundance and species richness.

```{r count-plotdata}
response.name <- 'total_abundance'

overzero <-  tidy.count %>% 
  filter(response %in% response.name & number > 0) 

equalzero <- tidy.count %>% 
  filter(response %in% response.name & number == 0)
```

Visualise the count data as a bubble plot. White bubbles represent zeroes and the size of blue bubbles represents abundance.

```{r visualise-count}
bubble.plot <- leaflet(data = tidy.count) %>%                     
  addTiles() %>%                                                    
  addProviderTiles('Esri.WorldImagery', group = "World Imagery") %>%
  addLayersControl(baseGroups = c("Open Street Map", "World Imagery"), options = layersControlOptions(collapsed = FALSE))

if (nrow(overzero)) {                                               
  bubble.plot <- bubble.plot %>%
    addCircleMarkers(data = overzero, lat = ~ latitude_dd, lng = ~ longitude_dd, radius = ~ number / 10, fillOpacity = 0.5, stroke = FALSE, label = ~ as.character(sample))}

if (nrow(equalzero)) {                                            
  bubble.plot <- bubble.plot %>%
    addCircleMarkers(data = equalzero, lat = ~ latitude_dd, lng = ~ longitude_dd, radius = 2, fillOpacity = 0.5, color = "white", stroke = FALSE, label = ~ as.character(sample))}
bubble.plot
```

# Visualise the fish length data
Filter the length data to only include one response. In this example the length metrics included are indicator species greater than and smaller than the length of maturity.

```{r length-plotdata}
response.name <- 'greater than Lm'

overzero <-  tidy.length %>% 
  dplyr::filter(response %in% response.name & number > 0) 

equalzero <- tidy.length %>% 
  dplyr::filter(response %in% response.name & number == 0)
```

Visualise the count data as a bubble plot. White bubbles represent zeroes and the size of blue bubbles represents abundance.

```{r visualise-length}
bubble.plot <- leaflet(data = tidy.length) %>%                     
  addTiles() %>%                                                    
  addProviderTiles('Esri.WorldImagery', group = "World Imagery") %>%
  addLayersControl(baseGroups = c("Open Street Map", "World Imagery"), options = layersControlOptions(collapsed = FALSE))

if (nrow(overzero)) {                                               
  bubble.plot <- bubble.plot %>%
    addCircleMarkers(data = overzero, lat = ~ latitude_dd, lng = ~ longitude_dd, radius = ~ number + 3, fillOpacity = 0.5, stroke = FALSE, label = ~ as.character(sample))}

if (nrow(equalzero)) {                                            
  bubble.plot <- bubble.plot %>%
    addCircleMarkers(data = equalzero, lat = ~ latitude_dd, lng = ~ longitude_dd, radius = 2, fillOpacity = 0.5, color = "white", stroke = FALSE, label = ~ as.character(sample))}
bubble.plot
```

# Save tidy datasets for modelling
Save count response data (total abundance and species richness).

```{r save-count}
saveRDS(tidy.count, file = here::here(paste0("r-workflows/data/tidy/", name, "_tidy-count.rds")))
```

Save length response data (indicator species greater than and smaller than the length of maturity).

```{r save-length}
saveRDS(tidy.length, file = here::here(paste0("r-workflows/data/tidy/", name, "_tidy-length.rds")))
```