---
title: "Generate spatial predictions of habitat using FSSgam"
author: "Claude Spencer & Brooke Gibbons"
date: "2023-11-13"
output:
  html_vignette:
    toc: true
    code_folding: hide
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Generate spatial predictions of habitat using FSSgam}
  %\VignetteEncoding{UTF-8}
---

This script takes the checked habitat data from the previous workflow steps, visualises the data and exports it into a format suitable for modelling. The exploratory visualisation of the data allows for trends and patterns in the raw data to be investigated.

```{r install-fssgam, include=FALSE}
devtools::install_github("beckyfisher/FSSgam_package", upgrade = "never")
```
# R setup
Load libraries.
```{r libraries, results='hide', warning=FALSE, message=FALSE}
library('remotes')
options(timeout=9999999)
# remotes::install_github("GlobalArchiveManual/CheckEM")
library(CheckEM)
library(tidyverse)
library(mgcv)
library(devtools)
library(FSSgam)
library(here)
library(ggplot2)
library(ggnewscale)
library(viridis)
library(terra)
library(sf)
library(patchwork)
```
Set the study name.
```{r name}
name <- 'example-bruv-workflow'
```
# Load data
Load the habitat point annotation data.
```{r raw-data}
dat <- readRDS(here::here(paste0("r-workflows/data/tidy/",
                      name, "_tidy-habitat.rds"))) %>%
  glimpse()
```
# Set up data for modelling
Set the predictor variables.
```{r predictors}
names(dat)
pred.vars <- c("mbdepth","roughness", "detrended",
               "slope", "tpi", "aspect", "tri")
```
Check for correlation of predictor variables and remove anything highly correlated (>0.95).
```{r correlations}
round(cor(dat[ , pred.vars]), 2)
```
Plot the individual predictors to assess if any transformations are necessary. We suggest to only use transformations when absolutely necessary. In the example dataset, most of the response variables have relatively balanced distributions, and therefor we have left them untransformed.
```{r transformations, warning = FALSE, message = FALSE}
plot_transformations(pred.vars = pred.vars, dat = dat)
```

Reset the predictor variables to remove any highly correlated variables and include any transformed variables.
```{r reset-predictors}
pred.vars <- c("depth_m","roughness", "detrended",
               "tpi", "aspect", "tri")
```

Check to make sure response variables have less than 80% zeroes. Full-subset GAM modelling will produce unreliable results if your data is too zero inflated.

```{r check-responses}
resp.vars.all = unique(as.character(dat$habitat))
resp.vars = character()
for(i in 1:length(resp.vars.all)){
  temp.dat = dat[which(dat$habitat == resp.vars.all[i]),]
  if(length(which(temp.dat$habitat == 0)) / nrow(temp.dat) < 0.8){
    resp.vars = c(resp.vars, resp.vars.all[i])}
}
resp.vars
```

Add the directory to save model outputs, and set up the R environment for model selection.

```{r environment-setup}
outdir    <- ("r-workflows/model-output/habitat/")
out.all   <- list()
var.imp   <- list()
```
# Run the full subset model selection process
This loop has been adapted from @beckyfisher/FSSgam, and examples and documentation is available on GitHub and in Fisher, R, Wilson, SK, Sin, TM, Lee, AC, Langlois, TJ. A simple function for full-subsets multiple regression in ecology with R. Ecol Evol. 2018; 8: 6104–6113. https://doi.org/10.1002/ece3.4134
```{r FSSgam, results='hide', warning=FALSE, message=FALSE}
for(i in 1:length(resp.vars)){
  print(resp.vars[i])
  use.dat <- dat[dat$habitat == resp.vars[i],]
  use.dat   <- as.data.frame(use.dat)
  Model1  <- gam(cbind(count, (total_points_annotated - count)) ~
                   s(mbdepth, bs = 'cr'),
                 family = binomial("logit"),  data = use.dat)

  model.set <- generate.model.set(use.dat = use.dat,
                                  test.fit = Model1,
                                  pred.vars.cont = pred.vars,
                                  cyclic.vars = c("aspect"),
                                  k = 5,
                                  cov.cutoff = 0.7
  )
  out.list <- fit.model.set(model.set,
                            max.models = 600,
                            parallel = T)
  names(out.list)

  out.list$failed.models
  mod.table <- out.list$mod.data.out
  mod.table <- mod.table[order(mod.table$AICc), ]
  mod.table$cumsum.wi <- cumsum(mod.table$wi.AICc)
  out.i     <- mod.table[which(mod.table$delta.AICc <= 2), ]
  out.all   <- c(out.all, list(out.i))
  var.imp   <- c(var.imp, list(out.list$variable.importance$aic$variable.weights.raw))

  for(m in 1:nrow(out.i)){
    best.model.name <- as.character(out.i$modname[m])

    png(file = here::here(paste(outdir, m, resp.vars[i], "mod_fits.png", sep = "")))
    if(best.model.name != "null"){
      par(mfrow = c(3, 1), mar = c(9, 4, 3, 1))
      best.model = out.list$success.models[[best.model.name]]
      plot(best.model, all.terms = T, pages = 1, residuals = T, pch = 16)
      mtext(side = 2, text = resp.vars[i], outer = F)}
    dev.off()
  }
}
```
Save the model fits and importance scores.
```{r save-gam-data}
names(out.all) <- resp.vars
names(var.imp) <- resp.vars
all.mod.fits <- list_rbind(out.all, names_to = "response")
all.var.imp  <- do.call("rbind",var.imp)
write.csv(all.mod.fits[ , -2], file = here::here(paste0(outdir, name, "_all.mod.fits.csv")))
write.csv(all.var.imp,         file = here::here(paste0(outdir, name, "_all.var.imp.csv")))
```
# Spatially predict the top model from the model selection process
Transform the habitat data into wide format for easy prediction.
```{r transform-habitat}
widedat <- dat %>%
  pivot_wider(values_from = "count", names_from = "habitat", values_fill = 0) %>%
  clean_names() %>%
  glimpse()
```
Load the raster of bathymetry data and derivatives.
```{r load-bathymetry, fig.width = 7, fig.height = 4}
preds  <- rast(here::here(paste0("r-workflows/data/spatial/rasters/",
                         name, "_bathymetry_derivatives.rds")))
plot(preds)
```

Transform the raster to a dataframe to predict onto.
```{r bathymetry-dataframe}
preddf <- as.data.frame(preds, xy = TRUE, na.rm = TRUE) %>%
  dplyr::mutate(depth = abs(mbdepth)) %>%
  clean_names() %>%
  glimpse()
```
Manually set the top model from the full subset model selection.
```{r set-models, fig.width = 7, fig.height = 6}
# Sessile invertebrates
m_inverts <- gam(cbind(sessile_invertebrates, total_points_annotated - sessile_invertebrates) ~
                 s(detrended,     k = 5, bs = "cr")  +
                 s(roughness, k = 5, bs = "cr") +
                 s(tpi, k = 5, bs = "cr"),
               data = widedat, method = "REML", family = binomial("logit"))
summary(m_inverts)
plot(m_inverts, pages = 1, residuals = T, cex = 5)

# # Rock
# m_rock <- gam(cbind(consolidated_hard, total_points_annotated - consolidated_hard) ~
#                    s(aspect,     k = 5, bs = "cc") +
#                    s(detrended, k = 5, bs = "cr") +
#                    s(tpi, k = 5, bs = "cr"),
#                  data = widedat, method = "REML", family = binomial("logit"))
# summary(m_rock)
# plot(m_rock, pages = 1, residuals = T, cex = 5)

# Sand
# m_sand <- gam(cbind(unconsolidated_soft, total_points_annotated - unconsolidated_soft) ~
#                 s(aspect,     k = 5, bs = "cc")  +
#                 s(detrended, k = 5, bs = "cr") +
#                 s(roughness, k = 5, bs = "cr"),
#               data = widedat, method = "REML", family = binomial("logit"))
# summary(m_sand)
# plot(m_sand, pages = 1, residuals = T, cex = 5)

# Seagrasses
m_seagrass <- gam(cbind(seagrasses, total_points_annotated - seagrasses) ~
                s(aspect,     k = 5, bs = "cc")  +
                s(detrended, k = 5, bs = "cr") +
                s(roughness, k = 5, bs = "cr"),
              data = widedat, method = "REML", family = binomial("logit"))
summary(m_seagrass)
plot(m_seagrass, pages = 1, residuals = T, cex = 5)

# Macroalgae
m_macro <- gam(cbind(macroalgae, total_points_annotated - macroalgae) ~
                 s(mbdepth,     k = 5, bs = "cr")  +
                 s(detrended, k = 5, bs = "cr") +
                 s(tpi, k = 5, bs = "cr"),
               data = widedat, method = "REML", family = binomial("logit"))
summary(m_macro)
plot(m_macro, pages = 1, residuals = T, cex = 5)

# Reef
m_reef <- gam(cbind(reef, total_points_annotated - reef) ~
                 s(aspect,     k = 5, bs = "cc")  +
                 s(detrended, k = 5, bs = "cr") +
                 s(roughness, k = 5, bs = "cr"),
               data = widedat, method = "REML", family = binomial("logit"))
summary(m_reef)
plot(m_reef, pages = 1, residuals = T, cex = 5)
```

# Predict, rasterise and plot habitat predictions
```{r spatial-prediction}
preddf <- cbind(preddf,
                "pmacro" = predict(m_macro, preddf, type = "response"),
                # "prock" = predict(m_rock, preddf, type = "response"),
                # "psand" = predict(m_sand, preddf, type = "response"),
                "pseagrass" = predict(m_seagrass, preddf, type = "response"),
                "pinverts" = predict(m_inverts, preddf, type = "response"),
                "preef" = predict(m_reef, preddf, type = "response")) %>%
  glimpse()
```

# Tidy and save the final dataset
Create a column for the 'dominant' habitat
```{r dominant-tag}
preddf$dom_tag <- apply(preddf %>% dplyr::select(pmacro, #prock, 
                                                 # psand, 
                                                 pseagrass, 
                                                 pinverts), 1,
                         FUN = function(x){names(which.max(x))})
glimpse(preddf)
```
Save final predictions
```{r save-final}
saveRDS(preddf, file = here::here(paste0("r-workflows/model-output/habitat/",
                              name, "_habitat-prediction.RDS")))
```
# Set up for plotting
Load marine park data. The dataset used here is the 2022 Collaborative Australian Protected Areas Database, which is available for free download from https://fed.dcceew.gov.au/datasets/782c02c691014efe8ffbd27445fe41d7_0/explore. Feel free to replace this shapefile with any suitable dataset that is available for your study area.
```{r marine-parks}
marine.parks <- st_read(here::here("r-workflows/data/spatial/shapefiles/Collaborative_Australian_Protected_Areas_Database_(CAPAD)_2022_-_Marine.shp")) %>%
  st_make_valid() %>%
  dplyr::mutate(ZONE_TYPE = str_replace_all(ZONE_TYPE,"\\s*\\([^\\)]+\\)", "")) %>%
  dplyr::filter(str_detect(ZONE_TYPE, "Sanctuary|National Park"), STATE %in% "WA") %>%
  st_transform(4326)
```
Set colours for habitat plotting.
```{r habitat-colours}
unique(preddf$dom_tag)
hab_fills <- scale_fill_manual(values = c(
  # "psand" = "wheat",
  "pinverts" = "plum",
  "pseagrass" = "forestgreen",
  "pmacro" = "darkgoldenrod4"
), name = "Habitat")
```
# Plot the dominant habitats
Build plot elements and display.
```{r plot-dominant, fig.width = 7, fig.height = 4}
ggplot() +
  geom_tile(data = preddf, aes(x, y, fill = dom_tag)) +
  hab_fills +
  new_scale_fill() +
  geom_sf(data = marine.parks, fill = NA, colour = "#7bbc63",
          size = 0.2, show.legend = F) +
  coord_sf(xlim = c(min(preddf$x), max(preddf$x)),
           ylim = c(min(preddf$y), max(preddf$y))) +
  labs(x = NULL, y = NULL, colour = NULL) +
  theme_minimal()
```

# Plot the individual habitat probabilities
Transform habitat predictions into long format for easy plotting with ggplot::facet_wrap.
```{r wide-habitat}
indclass <- preddf %>%
  pivot_longer(cols = starts_with("p"), names_to = "habitat",
               values_to = "Probability") %>%
  dplyr::mutate(habitat = case_when(habitat %in% "pinverts" ~ "Sessile invertebrates",
                                    habitat %in% "pmacro" ~ "Macroalgae",
                                    # habitat %in% "prock" ~ "Rock",
                                    # habitat %in% "psand" ~ "Sand",
                                    habitat %in% "pseagrass" ~ "Seagrass",
                                    habitat %in% "preef" ~ "Reef")) %>%
  glimpse()
```
Build plot elements for individual habitat probabilities.
```{r plot-individual, fig.width = 7, fig.height = 4}
ggplot() +
  geom_tile(data = indclass, aes(x, y, fill = Probability)) +
  scale_fill_viridis(option = "D", direction = -1) +
  new_scale_fill() +
  geom_sf(data = marine.parks, fill = NA, colour = "#7bbc63",
          size = 0.2, show.legend = F) +
  coord_sf(xlim = c(min(preddf$x), max(preddf$x)),
           ylim = c(min(preddf$y), max(preddf$y))) +
  labs(x = NULL, y = NULL, fill = "Probability") +
  theme_minimal() +
  facet_wrap(~habitat) +
  theme(axis.text.x = element_text(size = 7))
```