---
title: "Generate spatial layers for modelling"
author: "Claude Spencer & Brooke Gibbons"
date: "2023-11-13"
output:
  html_vignette:
    toc: true
    code_folding: hide
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Generate spatial layers for modelling}
  %\VignetteEncoding{UTF-8}
---
# R set up
Load the necessary libraries.
```{r libraries, results='hide', warning=FALSE, message=FALSE}
library(sf)
library(terra)
library(stars)
library(starsExtra)
library(tidyverse)
library(here)
```
Set the study name.
```{r study-name}
name <- 'example-bruv-workflow'
```
# Load bathymetry data and create bathymetry derivatives
The bathymetry data used in the example is a relatively low spatial resolution dataset downloaded from https://ecat.ga.gov.au/geonetwork/srv/eng/catalog.search#/metadata/148758. Feel free to replace with any suitable bathymetry in .tif format.
```{r bathy}
bathy <- rast(here::here("r-workflows/data/spatial/rasters/swc_ga250m.tif")) %>%
  clamp(upper = 0, lower = -250, values = F) %>%
  trim()
```
Calculate terrain derivatives.
```{r derivatives}
preds <- terrain(bathy, neighbors = 8,
                 v = c("slope", "aspect", "TPI", "TRI", "roughness"),           # Remove here as necessary
                 unit = "degrees")
```
Calculate detrended bathymetry.
```{r detrend-bathy}
zstar <- st_as_stars(bathy)                                                    
detre <- detrend(zstar, parallel = 8)                                           
detre <- as(object = detre, Class = "SpatRaster")                               
names(detre) <- c("detrended", "lineartrend")
```
Join bathymetry with terrain and detrended derivatives.
```{r join-rasters}
preds <- rast(list(bathy, preds, detre[[1]]))                                   
names(preds)[1] <- "mbdepth"
```
Save spatial-layers.
```{r save-layers}
preds_w <- wrap(preds)
saveRDS(preds_w, 
          file = here::here(paste0("r-workflows/data/spatial/rasters/", 
                        name, "_bathymetry_derivatives.rds")))
```
# Load metadata and extract site-specific derivatives
Read in the metadata.
```{r load-metadata}
metadata <- readRDS(here::here(paste0("r-workflows/data/tidy/", 
                            name, "_metadata.rds"))) %>%
  dplyr::mutate(longitude_dd = as.numeric(longitude_dd),
                latitude_dd = as.numeric(latitude_dd)) %>%
  glimpse()
```
Convert the metadata to a spatial file.
```{r vectorise-metadata}
metadata.vect <- vect(metadata, geom = c("longitude_dd", "latitude_dd"), crs = "epsg:4326")
```
Check to see if the metadata and bathymetry align correctly.
```{r check-alignment}
plot(preds[[1]])
plot(metadata.vect, add = T)
```
Add the latitude and longitude coordinates back into the metadata.
```{r add-coords}
tidy.metadata_t   <- as.data.frame(metadata.vect, geom = "XY") %>%
  left_join(metadata)
```
Extract the bathymetry derivatives at each sampling location.
```{r extract-derivatives}
metadata.bathy.derivatives   <- cbind(tidy.metadata_t, 
                                     terra::extract(preds, metadata.vect)) %>%
  dplyr::filter(!is.na(depth),                                                  
                !is.na(roughness)) %>%
  glimpse()
```
# Save the joined metadata and bathymetry derivatives
```{r save-metadata-bathy-derivatives}
saveRDS(metadata.bathy.derivatives, here::here(paste0("r-workflows/data/tidy/", name, "_metadata-bathymetry-derivatives.rds")))
```